---
title: "Using the medianeffect package for analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{median_effect_analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
csl: cell.csl
bibliography: medianeffect.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE, eval = FALSE}
suppressMessages(suppressWarnings(library(tidyverse)))
```

```{r}
library(medianeffect)
```


## Using the `medianeffect` R package for analysis

[@chouComputerizedQuantitationSynergism1994; @chouDerivationPropertiesMichaelisMenten1976]

```{r, eval = FALSE}
# Chou, T. C., R. J. Motzer, Y. Tong, and G. J. Bosl. 1994.
# “Computerized Quantitation of Synergism and Antagonism of Taxol, Topotecan, and Cisplatin against Human Teratocarcinoma Cell Growth: A Rational Approach to Clinical Protocol Design.”
# Journal of the National Cancer Institute 86 (20): 1517–24. https://doi.org/10.1093/jnci/86.20.1517.


# Paclitaxel
drug1 <- drug_effects(
  D = c(0.002, 0.004, 0.005, 0.01, 0.02),
  fa = c(0.429, 0.708, 0.761, 0.882, 0.932),
  name = "Paclitaxel",
  label = "Paclitaxel")
drug1


# Cis-Pt
drug2 <- drug_effects(
  D = c(0.05, 0.1, 0.2, 0.5, 1, 2),
  fa = c(0.055, 0.233, 0.301, 0.559, 0.821, 0.953),
  name = "Cisplatin",
  label = "cis-pt")
drug2


# Topotecan
drug3 <- drug_effects(
  D = c(0.01, 0.02, 0.05, 0.1, 0.2, 0.5),
  fa = c(0.069, 0.213, 0.373, 0.785, 0.94, 0.991),
  name = 'Topotecan',
  label = 'Topotecan'
)
drug3


# Paclitaxel:Cis-Pt Combo, ratio = 0.01
combo_1_2 <- drug_effects(
  D = c(0.001, 0.002, 0.005, 0.01) + c(0.1, 0.2, 0.5, 1),
  fa = c(0.45, 0.701, 0.910, 0.968),
  name = 'Paclitaxel - Cisplatin',
  label = 'combo_1_2',
  ratio = c(0.001, 0.1))
combo_1_2

# Create a 'fake' non-constant ratio object for testing
ncr_combo <- ncr_drug_effects(
  c(0.001, 0.002, 0.005, 0.01),
  c(0.1, 0.2, 0.5, 1),
  fa = c(0.45, 0.701, 0.910, 0.968),
  name='"NCR" Paclitaxel - Cisplatin',
  label = 'ncr_combo_1_2'
)
calc_ci(combo_1_2, drug1, drug2)
ncr_calc_ci(ncr_combo, drug1, drug2)

# Cis-Pt:Topotecan Combo, ratio = 0.1
combo_2_3 <- drug_effects(
  D = c(0.05, 0.1, 0.2, 0.5, 1) + c(0.005, 0.01, 0.02, 0.05, 0.1),
  fa = c(0.304, 0.413, 0.675, 0.924, 0.977),
  name = 'Cisplatin - Topotecan',
  label = 'combo_2_3',
  ratio = c(0.1, 0.01))
combo_2_3


# Paclitaxel:Topotecan Combo, ratio = 0.1
combo_1_3 <- drug_effects(
  D = c(0.001, 0.002, 0.005, 0.01) + c(0.01, 0.02, 0.05, 0.1),
  fa = c(0.274, 0.579, 0.901, 0.965),
  name = 'Paclitaxel - Topotecan',
  label = 'combo_1_3',
  ratio = c(0.01, 0.1))
combo_1_3


# Paclitaxel:Cis-Pt:Topotecan Combo, ratio = 0.001 : 0.1 : 0.01
combo_1_2_3 <- drug_effects(
  D = c(0.001, 0.002, 0.003, 0.005) + c(0.1, 0.2, 0.3, 0.5) + c(0.01, 0.02, 0.03, 0.05),
  fa = c(0.456, 0.806, 0.947, 0.995),
  name = 'Paclitaxel - Cisplatin - Topotecan',
  label = 'combo_1_2_3',
  ratio = c(0.001, 0.1, 0.01))
combo_1_2_3

median_effect_plot(drug1, drug2, drug3, combo_1_2, combo_1_3, combo_2_3, combo_1_2_3)

median_effect_plot(drug1, drug2, combo_1_2)

dose_effect_plot(drug1, drug2, drug3, combo_1_2, combo_1_3, combo_2_3,combo_1_2_3)
dose_effect_plot(drug1, drug2, drug3, combo_1_2, combo_1_3, combo_2_3,combo_1_2_3) +
  coord_cartesian(xlim = c(0, 2.1))

fa_ci_plot(combo_1_2_3, drug1, drug2, drug3)

fa_dri_plot(combo_1_2_3, drug1, drug2, drug3)
fa_dri_plot(combo_1_2_3, drug1, drug2, drug3) + scale_y_continuous(trans='log10')

drug1

drug2

drug3

combo_1_2
calc_ci(combo_1_2, drug1, drug2)

combo_2_3
calc_ci(combo_2_3, drug2, drug3)

combo_1_3
calc_ci(combo_1_3, drug1, drug3)

combo_1_2_3
calc_ci(combo_1_2, drug1, drug2, fa = c(0.5, 0.75, 0.9, 0.95))

calc_ci(combo_1_2_3, drug1, drug2, drug3)
calc_ci(combo_1_2_3, drug1, drug2, drug3, fa = c(0.5, 0.75, 0.9, 0.95))

calc_dri(combo_1_2_3, drug1, drug2, drug3, fa = c(0.5, 0.75, 0.9, 0.95))
```


## Appendix: Theoretical background

### Median effect equation

The median effect equation relates dose (D) to fraction affected (fa).

$$\frac{f_a}{f_u} = \left(\frac{D}{D_m}\right)^m$$

… where fraction affected plus fraction unaffected equals one, so: $f_u = 1 - f_a$

#### Dm parameter

At a dose of $D = D_m$, the fraction affected will equal 50%; hence "median effect."

#### m parameter

The $m$ parameter describes the shape of the dose-effect relationship:

* $m = 1$: hyperbolic
* $m > 1$: sigmoidal
* $m < 1$: negative (flat) sigmoidal

### Linearizing by log transformation

$$log\left(\frac{f_a}{1 - f_a}\right) = m~log(D) - m~log(D_m)$$

Fitting a linear regression of $log\left(\frac{f_a}{1 - f_a}\right)$ versus $log(D)$ allows determination of the $m$ and $D_m$ parameters.

### Solving for dose, D

$$D = D_m\left[\frac{f_a}{1 - f_a}\right]^{1/m}$$

### Solving for fraction affected, fa

$$f_a = \frac{1}{1 + \left(\frac{D_m}{D}\right)^m}$$

### Combination Index (CI)

$$CI = \sum_{i=1}^{n} \frac{(D)_i}{(D_x)_i}$$

… where for a given fraction affected $f_a$,

* $(D)_i$ is the dose of drug _i_ within the combination of _n_ drugs which together results in the fraction affected
* $(D_x)_i$ is the dose of drug _i_ alone which results in the fraction affected = $f_a$

The combination index describes the degree of synergism / antagonism, at a given fraction affected:

* CI = 1 indicates additive effect
* CI < 1 indicates synergism
* CI > 1 indicates antangonism

### Dose reduction index (DRI)

$$DRI_i = \frac{(D_x)_i}{(D)_i}$$

The reciprocal of each of the terms of the combination index (CI) reflects the fold-reduction (dose reduction index, DRI) of drug _i_ within the drug combination, at a given fractional affect $f_a$.

## References
